A Mechanised Library for Generic (Co)Programming

### 2024-02-22 paper/paper.tex
Initial algebras

Algebras on container extensions are defined as expected. The carrier of the
algebra is \mintinline{Coq}{A}, and the operations of the algebra are
morphisms of the following type:
\begin{minted}{Coq}
Notation Alg F A := (App F A ~> A).
\end{minted}

There are \emph{two} initial algebras that are, as expected, unique up to
isomorphism. We show one such initial algebra here, and the second in the
following subsection, plus a proof that they are isomorphic:

\begin{minted}{Coq}
Inductive LFix  : Type := LFix_in { LFix_out : App F LFix }.
\end{minted}
This is why we could not define equality of container extensions generically:
the contained types may not have (yet) an associated equivalence. We could have
overcame this circularity by defining simultaneously LFix and its associated
equivalence relation. But that would complicate definitions unnecessarily.
\dcas{I know ... I need to rewrite/expand on this}

We provide a custom induction schema for initial algebras:

\begin{minted}{Coq}
Lemma LFix_ind [P : LFix -> Prop]
  : (forall x : App F LFix,
        (forall e : Pos (shape x), P (cont x e)) ->
        P (LFix_in x))
    -> forall l : LFix, P l.
\end{minted}

Equality of initial algebras of container extensions needs to be defined
recursively and extensionally:
\begin{minted}{Coq}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{minted}

The unique homomorphism from an initial algebra to other algebras, i.e. the catamorphism, is defined as a
regular recursive function:
\begin{minted}{Coq}
Definition cata_f `{eA : equiv A} (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkCont sx (fun e => f (kx e)))
  end.
\end{minted}

>>>>

Algebras on container extensions are defined as expected. The carrier of the
algebra is \mintinline{Coq}{A}, and the operations of the algebra are morphisms
of the following type:
\begin{minted}{Coq}
Notation Alg F A := (App F A ~> A).
\end{minted}
Given a container extension \mintinline{Coq}{F}, we define \emph{two}
(isomorphic) initial algebras. We will show the need for the second initial
algebra in the following section. Our fist initial algebra is defined as the
straightforward least fixpoint of the container extension \mintinline{Coq}{F}:
\begin{minted}{Coq}
Inductive LFix  : Type := LFix_in { LFix_out : App F LFix }.
\end{minted}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words,
we would need to provide an equivalence for \mintinline{Coq}{LFix}, but the
only way to do this is to define such equivalence simultaneously to the
definition for  \mintinline{Coq}{LFix}.  We could have overcame this
circularity by defining  \mintinline{Coq}{LFix} independently of container
extensions, and define later the necessary
\mintinline{Coq}{in}/\mintinline{Coq}{out} functions to convert between
\mintinline{Coq}{LFix} and \mintinline{Coq}{App}. However, this extra layer
interferes with code extraction, in that it requires destructing/constructing
subsequent \mintinline{Coq}{LFix} and  \mintinline{Coq}{App}, which Coq is not
always able to inline.

We provide a custom induction scheme for initial algebras:
\begin{minted}{Coq}
Lemma LFix_ind [P : LFix -> Prop]
  : (forall x : App F LFix,
        (forall e : Pos (shape x), P (cont x e)) ->
        P (LFix_in x))
    -> forall l : LFix, P l.
\end{minted}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \mintinline{Coq}{AppR} in
Section~\ref{sec:TODO}.
\begin{minted}{Coq}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{minted}
From such initial algebras, we define \emph{catamorphisms} by recursively
traversing their structure. As usual, catamorphisms take an
\mintinline{Coq}{F}-algebra with carrier \mintinline{Coq}{A}, and produces a
function from \mintinline{Coq}{LFix F} to \mintinline{Coq}{A}.
\begin{minted}{Coq}
Definition cata_f `{equiv A} (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkCont sx (fun e => f (kx e)))
  end.
\end{minted}
